---
title: "Mortality delay analysis"
author: "Janet Rosenbaum"
date: "`r format(Sys.Date())`"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 2
    toc_float: true
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Define variables

## Libraries
```{r, message=F}
library(plm)
library(car)
library(lmtest)
library(mosaic)
library(lme4)
library(pbkrtest)
library(gee)
library(MASS)
library(nlme)
library(performance)
```

library(tidyr)
library(taRifx)
library(faraway)
library(robustlmm)
library(ggsci)
library(scales)

```{r, message=F, eval=F}
library(rstanarm)
library(bayesplot)
theme_set(bayesplot::theme_default())
```

## Variables

```{r}
setwd("~/OneDrive - Downstate Medical Center/projects/Coronavirus data/Mortality delay models")
MD=read.csv("weekly.delay.breakdown.by.state.03232021.csv", header=T)
GDP=read.csv("stategdp.csv", header=T)
```


```{r}
MD$delay=MD$Weeks.Delayed
MD$delayc=MD$delay- mean(MD$delay, na.rm=T)
MD$deathsphm=MD$Deaths.per.Million.of.State.Population/100
MD$logdeathsphm=log(MD$deathsphm)
MD$deathspm=MD$Deaths.per.Million.of.State.Population
MD$deaths=MD$Deaths..as.recorded.in.last.released.dataset.
MD$deathsplogpop=MD$deaths/log(MD$State.Population..in.millions.*10^6)
MD$logdeathsplogpop=log(MD$deaths)/log(MD$State.Population..in.millions.*10^6)
MD$logdeathspm = log(MD$Deaths.per.Million.of.State.Population)
```

```{r}
MDg=merge(MD, GDP, by.x="State", by.y="state")
MD=MDg
```

13 week completeness in 2017 Percent of death certificates available within 13 weeks.

Create variable for EDRS electronic death registration system
```{r}
MD$EDRS=2
MD$EDRS[MD$State=="Arkansas" | MD$State=="Colorado" | MD$State=="Maryland" | MD$State == "Michigan"| MD$State=="Mississippi" | MD$State=="New York"  | MD$State == "Pennsylvania"| MD$State == "Tennessee"| MD$State=="Virginia"]=1
MD$EDRS[MD$State=="Rhode Island" | MD$State=="Connecticut" | MD$State=="North Carolina" | MD$State == "West Virginia"]=0
MD$EDRSf=factor(MD$EDRS, levels=c(2, 1, 0), labels=c("Full", "Partial", "No"))

MD$EDRSno=MD$EDRSf=="No"
```

```{r}
MD$datef=with(MD, as.Date(Date))
MD$datef.c=(MD$datef-min(MD$datef))/7
MD$coroneronly=MD$Systems=="County/district-based coroner offices"
MD$countyDI=MD$Systems=="County/district-based coroner offices" | 
        MD$Systems=="County-based mixture of medical examiner and coroner offices" | 
  MD$Systems=="County/district-based medical examiner offices"
```

```{r}
MD$April=with(MD, datef=="2020-04-04" | datef=="2020-04-11" | datef=="2020-04-18" |
                datef=="2020-04-25")
MD$May=with(MD, datef=="2020-05-01" | datef=="2020-05-08" | datef=="2020-05-15" |
                datef=="2020-05-22" | datef=="2020-05-29")
MD$June=with(MD, datef=="2020-06-05" | datef=="2020-06-12" | datef=="2020-06-19" |
                datef=="2020-06-26")
MD$July=with(MD, datef=="2020-07-03" | datef=="2020-07-10" | datef=="2020-07-17" |
                datef=="2020-07-24"| datef=="2020-07-31")
MD$August=with(MD, datef=="2020-08-07" | datef=="2020-08-14" | datef=="2020-08-21" |
                datef=="2020-08-28")
```

```{r}
MD$Systems=with(MD, factor(Systems, labels=c("Centralized ME\nn=17", "County mixture\nn=14", "County coroner\nn=14", "County ME\nn=7")))
```


# Models

```{r}
femodel2=glmer(delay ~ EDRSf + deathsphm + (1 | State), data=MD, family=poisson)
femodel2bin=glmer(delay ~ EDRSno + deathsphm + (1 | State), data=MD, family=poisson)
summary(femodel2)
```

```{r}
durbinWatsonTest(residuals(femodel2), simulate=T, reps=1000, method="resample")
```

Autocorrelation --- AR as in ARIMA
```{r}
pacf(residuals(femodel2))
```

Moving averages: MA 
```{r}
acf(residuals(femodel2))
```

Try a model with log deaths (a reviewer suggestion) --- doesn't converge with log deaths per million, but it does converge with log deaths per log pop and Durbin Watson is worse.

```{r}
femodel2b=glmer(delay ~ EDRSf + logdeathsplogpop + (1 | State), data=MD, family=poisson)
durbinWatsonTest(residuals(femodel2b), simulate=T, reps=1000, method="resample")
```

```{r}
summary(femodel2b)
```
```{r}
exp(fixef(femodel2b))
exp(confint(femodel2b))
```


Try a model with lag deaths  and Durbin Watson is unchanged

```{r}
femodel2c=glmer(delay ~ EDRSf +  lag(deathsphm, 1) + (1 | State), data=MD, family=poisson)
durbinWatsonTest(residuals(femodel2c), simulate=T, reps=1000, method="resample")
```

Breusch-Godfrey.
```{r}
bgtest(delay ~ EDRSf + deathsphm, data=MD)
```

Serial correlation using the Panel data also.
```{r}
E <- pdata.frame(MD, index=c("State","datef"))
femodel2plm=plm(delay ~ EDRSf + deathsphm , data=E)
pbgtest(femodel2plm)
```


glmer does not have a correlation argument, but glmmPQL does have a correlation argument. glmmPQL is not optimized to deal with crossed random effects but the results of both functions are very often identical. 

```{r}
femodel2p=glmmPQL(fixed = delay ~ EDRSf + deathsphm, random=~ 1 | State, data=MD, family=poisson, correlation=corARMA(p=2, q=2))
summary(femodel2p)
```


```{r}
cbind(exp(1.42), exp(1.42-2*0.06), exp(1.42+2*0.06))
cbind(exp(-0.11), exp(-0.11-2*0.12), exp(-0.11+2*0.12))
cbind(exp(0.65), exp(0.65-2*0.17), exp(0.65+2*0.17))
cbind(exp(0.10), exp(0.10-2*0.01), exp(0.10+2*0.01))

```


```{r}
pacf(residuals(femodel2p))
```

```{r}
acf(residuals(femodel2p))
```

```{r}
plot(residuals(femodel2p))
```

```{r}
library(fitARMA)
library(astsa)
ar_ <- FitARMA(residuals(femodel2p), order = c(2, 0, 2), demean = FALSE)
acf2(residuals(ar_))
```


```{r}
acf(residuals(femodel2p))
```

```{r}
plot(femodel2p$residuals)
```

```{r}
check_autocorrelation(femodel2p)
```
```{r}
is.vector(residuals(femodel2p))
```

```{r}
durbinWatsonTest(as.vector(residuals(femodel2p)), simulate=T, reps=1000, method="resample")
```

